Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
more tables, more discussion

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2\%\xspace}
\newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{39210391\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....

@@

%
%
%
%
\subsection{Evaluation of Types from Traces}
\label{subsec:eval}

@@

@@

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
This number is low, and suggests that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

We breakdown the \PERCFAILEDASSERTIONS of failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${bf int}[]$, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
This is promising, as by-and-large it appears that \typetracer inferred argument type signatures which were not violated during testing.
But what do these failing assertions look like?
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.

The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
\AT{I need to touch up the discussion of this table.}

% While the raw number of failing checks is interesting, we were also curious to see if certain kinds of type checks were overrepresented in the data.
% A breakdown of the top failing type checks by kind can be found in Table~\ref{tbl:fail-breakdown-2}.
% % Note the last row in the table: this is a check that would succeed in a proper type system with subtyping for class types, as data frames are subclasses of tibbles (\code{tbl}), and we will revisit these scenarios in future work.

@@

>>>>

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2\%\xspace}
\newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}

@@

%
%
%
%
\subsection{RQ2: How good are type signatures generated from traces?}
\label{subsec:eval}

@@

%
%
\subsubsection{Overall Results}

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
These S3 functions represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

@@

We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${bf int}[]$, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
\AT{More on these functions?}

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.

The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.

@@

%
%
\subsubsection{Patterns in Failed Contract Assertions}

While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.

We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
